---
title: "ontoProc2 -- leveraging semantic SQL for ontology analysis in Bioconductor"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{ontoProc2 -- leveraging INCAtools semantic SQL for ontology analysis in Bioconductor}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

The ontoProc2 package has two aims:

- to give convenient
access to the ontologies that are transformed to "semantic
SQL" in the [INCAtools semantic sql project](https://github.com/INCATools/semantic-sql);
- to simplify operations that have been available in [ontoProc](https://git.bioconductor.org/packages/ontoProc), which will be deprecated in 2027.

# Acquiring ontologies

## Make a connection

The best way to work with an ontology in this system is
to use `semsql_connect`.  The `ontology` argument will
be a short string that the INCAtools project uses as part
of the filename for the ontology.  For Gene Ontology, the
string is "go".

```{r doini, message=FALSE}
library(ontoProc2)
goss = semsql_connect(ontology="go")
goss
```
## Make a report

The `report` method provides details.
```{r lkrep1}
report(goss)
```

## Probe the back end

The back end is SQLite.  We can enumerate the
tables available:
```{r lktbs}
library(dplyr)
library(DBI)
allt = dbListTables(goss@con)
length(allt)
head(allt)
```

## Use tidy methods

Individual tables are readily accessible.

```{r lkti,message=FALSE}
library(DT)
tbl(goss@con, "statements")
tbl(goss@con, "statements") |> head(20) |> as.data.frame() |> datatable()
```

To investigate the ontology, searching through RDF labels is a natural approach.

```{r dosrc}
search_labels(goss, "apoptosis") |> head() |> datatable()
```

Additional filtering could be useful here to focus on GO terms.  The `_riog...`
labels have special roles in RDF inference, and this will be addressed in 
vignettes to be added in the future.

Let's improve the query:
```{r dosrc2}
search_labels(goss, "apoptosis") |> filter(grepl("^GO:", subject)) |> head() |> datatable()
```
Clearly it will be valuable to filter away obsolete terms.  We will investigate
the use of edge tables to accomplish this in a future vignette.


# Transformation to ontology_index instances

The [ontologyX suite](https://academic.oup.com/bioinformatics/article/33/7/1104/2843897) of
Daniel Greene and colleagues provides very convenient ontology handling functions.
We can transform the SQLite data to this format.  We'll illustrate with cell ontology.

```{r lkoi}
clss = semsql_connect(ontology="cl")
cloi = semsql_to_oi(clss@con)
cloi
```

A convenience function assists with visualizations:

```{r dopl}
onto_plot2(cloi, c("CL:0000624", "CL:0000492", "CL:0000793", "CL:0000803"))
```

# Session information

```{r lksess}
sessionInfo()
```

